Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2021-11-03 move_schedules.tex
Arxiv version

\documentclass[review]{siamart190516}

% Journal name: SICOMP
%\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{scalefnt}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{algorithm}
%\usepackage{algorithm2e}
%\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\usepackage{bbm}
\usepackage{graphicx}

@@

%\newtheorem{theorem}{Theorem}
%\newtheorem{definition}{Definition}
%\newtheorem{proposition}{Proposition}
%\newtheorem{lemma}{Lemma}
%\newtheorem{corollary}{Corollary}
% \newcommand{\CommentX}[1]{\unskip~\%~#1~\%}

\newsiamremark{remark}{Remark}

@@

% Sets running headers as well as PDF title and authors
\headers{Fast Persistence Computations}{M. Piekenbrock and J. Perea}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}
}

@@

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={Fast Persistence Computations},
  pdfauthor={Matt Piekenbrock and Jose A. Perea}
}
\fi

% ------  Begin document ------- 
% https://github.com/peekxc/dart
\begin{document} 
\maketitle

@@

% Keywords and MSC codes must accompany all articles. A list of the subject classifications can be accessed or searched online in the Annual Index of Mathematical Reviews.
% REQUIRED
\begin{keywords}
  Computational topology, Persistent homology, Topological data analysis
\end{keywords}

% REQUIRED
% See: https://mathscinet.ams.org/mathscinet/msc/msc2020.html?t=68T09&btn=Current
\begin{AMS}
  68T09, 55N31, 62R40
\end{AMS}

@@

 \begin{enumerate}[leftmargin=2\parindent, align=left, labelsep=-5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
	\item[\hspace{-1.5em}\textbf{ Decomposition Invariants:}]

@@

 	\item[I2] $V$ is full-rank upper-triangular, and $R$ is \emph{reduced}: if its $i$-th and $j$-th columns are nonzero, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[leftmargin=2\parindent, align=left, labelsep=-5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in either $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid  during the movement of $\sigma_i $ to $ \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times.
By comparison, if a row-oriented sparse matrix implementation is used, then a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

@@

\begin{remark}
We note that while the pseudocode for \emph{MoveRight} is included in~\cite{busaryev2010tracking}, the one for \emph{MoveLeft} is not, and Algorithm~\ref{alg:ml} is the one we propose and leverage in this paper.
\end{remark}

>>>>

% \documentclass[review]{siamart190516} % for SICOMP 
\documentclass{article} % for arxiv 
\usepackage{arxiv}

%\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{url}
\usepackage{enumitem}
\usepackage{setspace}
\usepackage{scalefnt}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{algorithm}
%\usepackage{algorithm2e}
%\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\usepackage{bbm}
\usepackage{graphicx}

\pagestyle{plain} % arxiv only

@@

% For arxiv; comment out for SICOMP
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proof}{Proof}
% \newcommand{\CommentX}[1]{\unskip~\%~#1~\%}

%\newsiamremark{remark}{Remark}

@@

% Sets running headers as well as PDF title and authors
% \headers{Fast Persistence Computations}{M. Piekenbrock and J. Perea}

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{
	Move Schedules: Fast persistence computations in coarse dynamic settings
	\thanks{
	\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.
	}
}

@@

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={Move Schedules},
  pdfauthor={Matt Piekenbrock and Jose A. Perea}
}
\fi

% ------  Begin document ------- 
% https://github.com/peekxc/dart
\begin{document}

\maketitle

@@

% Keywords and MSC codes must accompany all articles. A list of the subject classifications can be accessed or searched online in the Annual Index of Mathematical Reviews.
% REQUIRED
\begin{keywords}
  {Computational topology, Persistent homology, Topological data analysis}
\end{keywords}

% REQUIRED
% See: https://mathscinet.ams.org/mathscinet/msc/msc2020.html?t=68T09&btn=Current
%\begin{AMS}
%  68T09, 55N31, 62R40
%\end{AMS}

@@

  \textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=2pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]

@@

 	\item[I2] $V$ is full-rank upper-triangular, and $R$ is \emph{reduced}: if its $i$-th and $j$-th columns are nonzero, \\ then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 

@@

Though they are similar, move operations confer a few computational advantages compared to vineyards. Informally, a move operation has two convenient properties:
\\	
 \begin{enumerate}[labelsep=5pt, topsep=0pt,itemsep=-0.25ex,parsep=1.2ex]
 	\item[M1:] Querying the non-zero status of entries in either $R$ or $V$ occurs once per move.
 	\item[M2:] $R = D V$ is \underline{not} guaranteed to be valid  during the movement of $\sigma_i $ to $ \sigma_j$.
 \end{enumerate} 
 \vspace*{1em}
First, consider property M1. Prior to applying any permutation $P$ to the decomposition, it is necessary to remove non-zero entries in $V$ which render $P^TVP$ non-upper triangular, to maintain invariant I2. 
If one uses the (augmented) sparse matrix representation described in~\cite{cohen2006vines}, checking whether these entries are non-zero takes $O(\log m)$ per column. Since moving a simplex from $i$ to $j$ using vineyards performs $\lvert i - j \rvert - 1$ transpositions, this query operation is required at most $O(\lvert i - j \rvert)$ times.
By comparison, if a row-oriented sparse matrix implementation is used, then a move operation can perform these non-zero entry checks in just one $O(m)$ pass, prior to performing any column operations.

@@

%\begin{remark}
%We note that while the pseudocode for \emph{MoveRight} is included in~\cite{busaryev2010tracking}, the one for \emph{MoveLeft} is not, and Algorithm~\ref{alg:ml} is the one we propose and leverage in this paper.
%\end{remark}
